---
title: "LD Exploration"
author: "K. Grinde"
date: "October 16, 2024"
format: 
  html:
    toc: true
    embed-resources: true
    code-fold: true
---

# Set Up

## Load Packages

```{r}
#| message: false
library(tidyverse)
library(MASS)
library(reshape2)
```

## Session Info

```{r}
#| collapse: true
#| code-fold: true
sessionInfo()
```



# Generating Toy Data

## Example 1: No LD

```{r}
N <- 100
P <- 10
MU <- rep(0, times = P)
SIGMA1 <- diag(1, nrow = P, ncol = P)

set.seed(1)
X1 <- mvrnorm(n = N, mu = MU, Sigma = SIGMA1)
```


Visualize the covariance matrix: 

```{r}
plot_corr <- function(X){
  # calculate empirical correlation
  cor_matrix <- cor(X)
  cor_matrix[lower.tri(cor_matrix)] <- NA
  cor_matrix <- cor_matrix %>% 
    melt() %>% 
    na.omit() %>% 
    rename(correlation = value)

  # Visualize the correlation for each pair of variables
  ggplot(cor_matrix, aes(x = Var1, y = Var2, fill = correlation)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(
      low = "blue", high = "red", mid = "white", 
      midpoint = 0, limit = c(-1,1)) +
    labs(x = "", y = "") +
    theme_minimal() + 
    theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
    coord_fixed()
}

plot_corr(X1)
```


## Example 2: Short LD Block

```{r}
SIGMA2 <- diag(1, nrow = P, ncol = P)
SIGMA2[4,5] <- SIGMA2[5,6] <- SIGMA2[6,7] <- 0.8
SIGMA2[5,4] <- SIGMA2[6,5] <- SIGMA2[7,6] <- 0.8
SIGMA2[4,6] <- SIGMA2[5,7] <- 0.6
SIGMA2[6,4] <- SIGMA2[7,5] <- 0.6
SIGMA2[4,7] <- 0.4
SIGMA2[7,4] <- 0.4

set.seed(1)
X2 <- mvrnorm(n = N, mu = MU, Sigma = SIGMA2)
```


Visualize the covariance matrix: 

```{r}
plot_corr(X2)
```

## Example 3: More SNPs

```{r}
P <- 50
MU3 <- rep(0, times = P)

SIGMA3 <- diag(1, nrow = P, ncol = P)
SIGMA3[24,25] <- SIGMA3[25,26] <- SIGMA3[26,27] <- SIGMA3[27,28] <- .8
SIGMA3[25,24] <- SIGMA3[26,25] <- SIGMA3[27,26] <- SIGMA3[28,27] <- .8
SIGMA3[24,26] <- SIGMA3[25,27] <- SIGMA3[26,28] <- 0.6
SIGMA3[26,24] <- SIGMA3[27,25] <- SIGMA3[28,26] <- 0.6
SIGMA3[24,27] <- SIGMA3[25,28] <- 0.4
SIGMA3[27,24] <- SIGMA3[28,25] <- 0.4


set.seed(1)
X3 <- mvrnorm(n = N, mu = MU3, Sigma = SIGMA3)
```


Visualize the covariance matrix: 

```{r}
plot_corr(X3)
```


## Example 4: Slower Decay

```{r}
P <- 10
SIGMA4 <- diag(1, nrow = P, ncol = P)
SIGMA4[4,5] <- SIGMA4[5,6] <- SIGMA4[6,7] <- 0.9
SIGMA4[5,4] <- SIGMA4[6,5] <- SIGMA4[7,6] <- 0.9
SIGMA4[4,6] <- SIGMA4[5,7] <- 0.8
SIGMA4[6,4] <- SIGMA4[7,5] <- 0.8
SIGMA4[4,7] <- 0.8
SIGMA4[7,4] <- 0.8

SIGMA4[3,4] <- SIGMA4[4,3] <- 0.4
SIGMA4[3,5] <- SIGMA4[5,3] <- 0.4
SIGMA4[3,6] <- SIGMA4[6,3] <- 0.2
SIGMA4[3,7] <- SIGMA4[7,3] <- 0.2

SIGMA4[2,3] <- SIGMA4[3,2] <- 0.3
#SIGMA4[2,4] <- SIGMA4[4,2] <- 0.3

SIGMA4[1,2] <- SIGMA4[2,1] <- 0.3
SIGMA4[9,10] <- SIGMA4[10,9] <- 0.3

set.seed(1)
X4 <- mvrnorm(n = N, mu = MU, Sigma = SIGMA4)
```


Visualize the covariance matrix: 

```{r}
plot_corr(X4)
```


## Next Steps

- add (way) more SNPs
- create two "chromosomes" / blocks that are unlinked but have decaying correlation within
- set it up so one of those two chromosomes has a much more extreme high LD block



# PCA

```{r}
run_pca <- function(X){
  pca <- prcomp(X, center = TRUE, scale = TRUE)
  loadings <- pca$rotation
  scores <- pca$x
  return(list(pca = pca, loadings = loadings, scores = scores))
}

plot_loadings <- function(X, npcs){
  pca <- run_pca(X)
  pca$loadings %>%
    as.data.frame() %>%
    mutate(index = seq_len(nrow(pca$loadings))) %>%
    melt(id="index") %>%
    filter(variable %in% paste0("PC",1:npcs)) %>%
    ggplot(aes(x = index, y = abs(value))) + 
    geom_point() + 
    facet_wrap(.~variable) +
    labs(xlab = 'SNP Number', ylab = 'Loadings') + 
    theme_bw()
}
```

```{r}
#| code-fold: false
plot_loadings(X1, 4)
```

```{r}
#| code-fold: false
plot_loadings(X2, 4)
```


```{r}
#| code-fold: false
plot_loadings(X3, 4)
```

```{r}
#| code-fold: false
plot_loadings(X4, 4)
```
